---
title: "Assignment 5: Water Quality in Lakes"
author: "Catherine Otero"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on water quality in lakes

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/fSe18vMhgzcjUKM39

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-02-22

## Setup

1. Verify your working directory is set to the R project file. Load the tidyverse, lubridate, and LAGOSNE packages. Set your ggplot theme (can be theme_classic or something else)
2. Load the LAGOSdata database and the trophic state index csv file we created in class.

```{r setup}
library(tidyverse)
library(lubridate)
library(LAGOSNE)

theme_set(theme_classic())

#set number of decimal points
options(scipen = 4)

# Load LAGOSNE data into R session
LAGOSdata <- lagosne_load()

#load trophic states csv

LAGOStrophic <- read.csv("./Data/Processed/LAGOStrophic.csv")

```

## Trophic State Index

3. Similar to the trophic.class column we created in class (determined from TSI.chl values), create two additional columns in the data frame that determine trophic class from TSI.secchi and TSI.tp (call these trophic.class.secchi and trophic.class.tp). 

```{r}
LAGOStrophic <- LAGOStrophic %>%
  mutate(trophic.class.secchi = case_when(TSI.secchi < 40 ~ "Oligotrophic",
  TSI.secchi >= 40 & TSI.secchi < 50 ~ "Mesotrophic",
  TSI.secchi >= 50 & TSI.secchi< 70 ~ "Eutrophic",
  TSI.secchi >= 70 ~ "Hypereutrophic"),
  trophic.class.tp = case_when(TSI.tp < 40 ~ "Oligotrophic",
  TSI.tp >= 40 & TSI.tp < 50 ~ "Mesotrophic",
  TSI.tp >= 50 & TSI.tp< 70 ~ "Eutrophic",
  TSI.tp >= 70 ~ "Hypereutrophic"))
  
```


4. How many observations fall into the four trophic state categories for the three metrics (trophic.class, trophic.class.secchi, trophic.class.tp)? Hint: `n()` function.

```{r}
#n() function gives the current group size. It is a context dependent expression so needs to be used within other functions like summarise and mutate
LAGOStrophic %>%
  group_by(trophic.class) %>%
  summarise(count = n())

# 1 Eutrophic      37457
# 2 Hypereutrophic 13234
# 3 Mesotrophic    13964
# 4 Oligotrophic    2762

LAGOStrophic %>%
  group_by(trophic.class.secchi) %>%
  summarise(count = n())

# 1 Eutrophic            25793
# 2 Hypereutrophic        4721
# 3 Mesotrophic          22344
# 4 Oligotrophic         14559

LAGOStrophic %>%
  group_by(trophic.class.tp) %>%
  summarise(count = n())

# 1 Eutrophic        22419
# 2 Hypereutrophic    6407
# 3 Mesotrophic      20607
# 4 Oligotrophic     17984
```

5. What proportion of total observations are considered eutrophic or hypereutrophic according to the three different metrics (trophic.class, trophic.class.secchi, trophic.class.tp)?

```{r}
LAGOStrophic %>%
  group_by(trophic.class) %>%
  summarise(count = n()) %>%
  mutate(prop = count/sum(count))

# Eutrophic 0.556
# Hypereutrophic 0.196

LAGOStrophic %>%
  group_by(trophic.class.secchi) %>%
  summarise(count = n()) %>%
  mutate(prop = count/sum(count))

# Eutrophic 0.383
# Hypereutrophic 0.0700

LAGOStrophic %>%
  group_by(trophic.class.tp) %>%
  summarise(count = n()) %>%
  mutate(prop = count/sum(count))

# Eutrophic 0.333
# Hypeeutrophic 0.0950

```

Which of these metrics is most conservative in its designation of eutrophic conditions? Why might this be?

> The measurement of chlorophyll a concentraation is the most conservative. Eutrophic is 56% of the samples and Hypereutrophic is 20% of the samples. Chlorophyll may be prominent in the lakes and causing the lakes to be murky.


## Nutrient Concentrations

6. Create a data frame that includes the columns lagoslakeid, sampledate, tn, tp, state, and state_name. Mutate this data frame to include sampleyear and samplemonth columns as well. Filter the data frame for May-September. Call this data frame LAGOSNandP. 

```{r}
LAGOSstate <- LAGOSdata$state
LAGOSnutrient <- LAGOSdata$epi_nutr
LAGOSlocus <- LAGOSdata$locus

LAGOSNandP <- LAGOSstate %>%
  left_join(., LAGOSlocus) %>%
  left_join(., LAGOSnutrient) %>%
  select(lagoslakeid, sampledate, tn, tp, state, state_name) %>%
  mutate(sampleyear = year(sampledate),
  samplemonth = month(sampledate)) %>%
  filter(samplemonth %in% c(5:9))
## Joining, by = "state_zoneid"
## Joining, by = "lagoslakeid"

```

7. Create two violin plots comparing TN and TP concentrations across states. Include a 50th percentile line inside the violins. Create a logged y axis and relabel axes.
```{r}
ggplot(LAGOSNandP, aes(x = state, y = tp)) +
  geom_violin(draw_quantiles = 0.50) +
  scale_y_log10() +
  labs(x = "State", y = expression("TP ("*mu*"g/L)"))

ggplot(LAGOSNandP, aes(x = state, y = tn)) +
  geom_violin(draw_quantiles = 0.50) +
  scale_y_log10() +
  labs(x = "State", y = expression("TN ("*mu*"g/L)"))
```

> TN: high: IA, OH Low: CT, ME, NH, VT

> TP: high: IA, IL, OH low: CT, ME, MI, NH

Which states have the largest and smallest concentration ranges?

> TN: large: IA, NY, RI small: NH, VT

> TP: large: CT small: CT, ME, MI, NH

8. Create two jitter plots comparing TN and TP concentrations across states, with samplemonth as the color. Choose a color palette other than the ggplot default.

```{r}
ggplot(LAGOSNandP, aes(x = state, y = tp, color = samplemonth)) +
  geom_jitter(alpha = 0.2, size = 0.5) +
  scale_y_log10() +
  labs(x = "State", y = expression("TP ("*mu*"g/L)")) +
  scale_color_viridis_c(option = "magma", begin = 0.2, end = 0.8, direction = -1)

ggplot(LAGOSNandP, aes(x = state, y = tn, color = samplemonth)) +
  geom_jitter(alpha = 0.2, size = 0.5) +
  scale_y_log10() +
  labs(x = "State", y = expression("TN ("*mu*"g/L)")) +
  scale_color_viridis_c(option = "magma", begin = 0.2, end = 0.8, direction = -1)
```

Which states have the most samples? How might this have impacted total ranges from #7?

> TN: IL, IN, MA, NH, NJ, VT
> TP: MA, NJ, OH

States with fewer samples have smaller ranges for TN and TP.


